Evolutionary rescue by aneuploidy in tumors exposed to anti-cancer drugs

### 2025-01-08 EvolutionaryRescue1.tex
changed align to equation + aligned
apparently align doesn't play well with line numbering

The cell population at time $t$ is represented by the triplet $\left(s_t,a_t,m_t\right)$. The following describes the events that may occur (right column), the rates at which they occur (middle column), and the effect these events have on the population (left column, see \Cref{figureAneuploidy}):

\begin{align*}
(+1,0,0)&:\quad \lambda_ss_t\left(1-u-v\right)\quad\left(\text{birth of sensitive cell}\right),\\
(-1,0,0)&:\quad \mu_ss_t\quad\left(\text{death of sensitive cell}\right),\\
(0,+1,0)&:\quad u\lambda_ss_t\quad\left(\text{sensitive cell divides and becomes aneuploid}\right),\\
(0,0,+1)&:\quad v\lambda_ss_t\quad\left(\text{sensitive cell divides and becomes mutant}\right),\\
(0,+1,0)&:\quad \lambda_aa_t\left(1-v\right)\quad\left(\text{birth of aneuploid cell}\right),\\
(0,-1,0)&:\quad \mu_aa_t\quad\left(\text{death of aneuploid cell}\right),\\
(0,0,+1)&:\quad v\lambda_aa_t\quad\left(\text{aneuploid cell divides and becomes mutant}\right),\\
(0,0,+1)&:\quad \lambda_mm_t\quad\left(\text{birth of mutant cell}\right),\\
(0,0,-1)&:\quad \mu_mm_t\quad\left(\text{death of mutant cell}\right).
\end{align*}
For the remaining of this paper, we assume that the division rates of sensitive and aneuploid cells can be written as $\lambda_ss_t\left(1-u-v\right)\approx \lambda_ss_t$ and $\lambda_aa_t\left(1-v\right)\approx\lambda_aa_t$ because $u,v\ll1$ (\Cref{table1}).
Each iteration of the simulation loop starts by computing the rates $\nu_k$ of each event $k$.
We then draw the time until the next event, $\Delta t$, from an exponential distribution whose rate parameter is the sum of the rates of all events, such that $\Delta t \sim \text{Exp}(\sum_j \nu_j)$.
Then, we randomly determine which event occurred, where the probability for event $k$ is $p_k = \nu_k/\sum_j \nu_j$.
Finally, we update the number of cells of each genotype according to the event that occurred and update the time from $t$ to $t+\Delta t$.
We repeat these iterations until either the population becomes extinct (the number of cells of all genotypes is zero) or the number of mutant cells is high enough so that their extinction probability is $<0.1\%$, that is, until
\begin{equation*}
m_t > \left\lfloor\frac{3\log{10}}{\log{\left(\lambda_m / \mu_m\right)}}\right\rfloor + 1 ,
\end{equation*}
which we obtain by solving $1-(1-p_m)^{m_t}=0.999$ for $m_t$ with $p_m=\Delta_m/\lambda_m$ as the probability that a single mutant escapes stochastic extinction (Appendix A).

@@

In our analysis, we assume that cells from the initial population divide and die independently of each other. However, these cells will compete for resources. We assume this competition can be ignored because the drug will cause the cell density to rapidly drop below the carrying capacity where competition is important.
To test this assumption, we simulate a logistic growth model, with division and death rates given by
\begin{align*}
\lambda_s' &= \lambda_s , \\
\mu_s' &= \mu_s ,\\
\lambda_a' &= \lambda_a ,\\ 
\mu_a' &= \mu_a + \lambda_a\frac{w+a+m}{K} ,\\
\lambda_m' &= \lambda_m ,\\ 
\mu_m' &= \mu_m + \lambda_m\frac{w+a+m}{K} ,
\end{align*}
where $K$ is the tumor carrying capacity. 
The effective carrying capacity in this model is $K_e=K\Delta_a/\lambda_a\approx10^6$ for $K=10^8, \lambda_a=0.0901,\mu_a=0.09$, where we define the effective carrying capacity to be the population size at which the aneuploid division rate is equal to the aneuploid death rate.

>>>>

The cell population at time $t$ is represented by the triplet $\left(s_t,a_t,m_t\right)$. The following describes the events that may occur (right column), the rates at which they occur (middle column), and the effect these events have on the population (left column, see \Cref{figureAneuploidy}):
\[
\begin{aligned}
(+1,0,0)&:\quad \lambda_ss_t\left(1-u-v\right)\quad\left(\text{birth of sensitive cell}\right),\\
(-1,0,0)&:\quad \mu_ss_t\quad\left(\text{death of sensitive cell}\right),\\
(0,+1,0)&:\quad u\lambda_ss_t\quad\left(\text{sensitive cell divides and becomes aneuploid}\right),\\
(0,0,+1)&:\quad v\lambda_ss_t\quad\left(\text{sensitive cell divides and becomes mutant}\right),\\
(0,+1,0)&:\quad \lambda_aa_t\left(1-v\right)\quad\left(\text{birth of aneuploid cell}\right),\\
(0,-1,0)&:\quad \mu_aa_t\quad\left(\text{death of aneuploid cell}\right),\\
(0,0,+1)&:\quad v\lambda_aa_t\quad\left(\text{aneuploid cell divides and becomes mutant}\right),\\
(0,0,+1)&:\quad \lambda_mm_t\quad\left(\text{birth of mutant cell}\right),\\
(0,0,-1)&:\quad \mu_mm_t\quad\left(\text{death of mutant cell}\right).
\end{aligned}
\]

For the remainder of this paper, we assume that the division rates of sensitive and aneuploid cells can be written as $\lambda_ss_t\left(1-u-v\right)\approx \lambda_ss_t$ and $\lambda_aa_t\left(1-v\right)\approx\lambda_aa_t$ because $u,v\ll1$ (\Cref{table1}).
Each iteration of the simulation loop starts by computing the rates $\nu_k$ of each event $k$.
We then draw the time until the next event, $\Delta t$, from an exponential distribution whose rate parameter is the sum of the rates of all events, such that $\Delta t \sim \text{Exp}(\sum_j \nu_j)$.
Then, we randomly determine which event occurred, where the probability for event $k$ is $p_k = \nu_k/\sum_j \nu_j$.
Finally, we update the number of cells of each genotype according to the event that occurred and update the time from $t$ to $t+\Delta t$.
We repeat these iterations until either the population becomes extinct (the number of cells of all genotypes is zero) or the number of mutant cells is high enough so that their extinction probability is $<0.1\%$, that is, until
\begin{equation*}
m_t > \left\lfloor\frac{3\log{10}}{\log{\left(\lambda_m / \mu_m\right)}}\right\rfloor + 1 ,
\end{equation*}
which we obtain by solving $1-(1-p_m)^{m_t}=0.999$ for $m_t$ with $p_m=\Delta_m/\lambda_m$ as the probability that a single mutant escapes stochastic extinction (Appendix A).

@@

In our analysis, we assume that cells from the initial population divide and die independently of each other. However, these cells will compete for resources. We assume this competition can be ignored because the drug will cause the cell density to rapidly drop below the carrying capacity where competition is important.
To test this assumption, we simulate a logistic growth model, with division and death rates given by
\[
\begin{aligned}
\lambda_s' &= \lambda_s , \\
\mu_s' &= \mu_s ,\\
\lambda_a' &= \lambda_a ,\\ 
\mu_a' &= \mu_a + \lambda_a\frac{w+a+m}{K} ,\\
\lambda_m' &= \lambda_m ,\\ 
\mu_m' &= \mu_m + \lambda_m\frac{w+a+m}{K} ,
\end{aligned}
\]
where $K$ is the tumor carrying capacity. 
The effective carrying capacity in this model is $K_e=K\Delta_a/\lambda_a\approx10^6$ for $K=10^8, \lambda_a=0.0901,\mu_a=0.09$, where we define the effective carrying capacity to be the population size at which the aneuploid division rate is equal to the aneuploid death rate.
